Parallel Adaptive Importance Sampling

### 2015-08-14 PAIS.tex
Cleared out old figures
Added section on calculating speed up

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins}
\setlength{\parindent}{0in}
\usepackage{amsmath,amsfonts,amsbsy}
\newcommand{\RARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
\newcommand{\RARRlong}[3]{#1
  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
  \longrightarrow}^{#3}\; #2}
\newcommand{\LARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
\newcommand{\bX}{{\bf X}}
\newcommand{\vecx}{{\mathbf x}}
\newcommand{\vecy}{{\mathbf y}}
\newcommand{\vecz}{{\mathbf z}}
\newcommand{\vecq}{{\mathbf q}}
\newcommand{\bs}{{\mathbf s}}
\newcommand{\vecr}{{\mathbf r}}
\newcommand{\vecX}{{\mathbf X}}
\newcommand{\vecv}{{\mathbf v}}
\newcommand{\tick}{\ding{52}}
\newcommand{\cross}{\ding{54}}
\newcommand{\vecn}{{\mathbf n}}
\newcommand{\vecp}{{\mathbf p}}
\newcommand{\cT}{{\mathcal T}}
\newcommand{\dt}{{\mbox{d}t}}
\newcommand{\dx}{{\mbox{d} \vecx}}
\newcommand{\boldnu}{{\boldsymbol \nu}}
\newcommand{\er}{{\mathbb R}}
\renewcommand{\div}{{\rm div}}
\newcommand{\bnu}{{\bf \nu}}
\newcommand{\divergence}{\mathop{\mbox{div}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\FP}{P_{\rm{FP}}}
\newcommand{\ME}{P_{\rm{ME}}}
\newcommand{\MEs}{P_{\rm{ME}_{S}}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\X}{{\mathbf X}}
\newcommand{\Y}{{\mathbf Y}}
\newcommand{\W}{{\mathbf W}}
\newcommand{\data}{D}
\newcommand{\neff}{n_{\text{eff}}}
\newcommand{\E}{{\mathbb E}}
\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are commonly used and powerful
  family of numerical methods for sampling from complex probability
  distributions. As the applications that these methods are being
  applied to increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and with the development of
faster more powerful computers, have become ever more important in a
whole range of fields in statistics, science and engineering. However,
despite the advances in hardware that have made it possible to fully
characteries the posterior distributions of many problems, it remains
unfeasible to use MCMC methods for many more. In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

@@

Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments. To reach a desired tolerance, the PAIS-pCNL algorithm requires 40\% fewer iterations than the pCNL algorithm.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{"figures/pCNL2 opt delta"}
\includegraphics[width=0.45\textwidth]{"figures/PAIS pCNL2 opt delta"}
\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}. The setup is as in Section~\ref{sec:Implementation P1}.}
\label{fig:P2 opt delta}
\end{figure}

@@

Figure~\ref{fig:MH2 L2} shows that the PAIS-pCNL algorithm, converges to the posterior distribution faster than the pCNL algorithm. The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations. We see that to obtain a given tolerance level, the PAIS-pCNL algorithm requires 35\% fewer iterations than the pCNL algorithm.

@@

We can see that the adaptive algorithms compare closely with the
respective nonadaptive algorithms and the improvement PAIS offers
remains significant. The PAIS-pCNL algorithm requires 45\% fewer
iterations than the naively parallelised pCNL algorithm.

@@

\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_suction"}
\caption{This figure demonstrates the redistribution property of the PAIS algorithm. Initially there is one chain in the positive mode, and 49 chains in the negative mode.}
\label{fig:BM2_suction}
\end{center}
\end{figure}

@@

\begin{appendix}
\section{The adaptive PAIS algorithm}\label{Sec:App}

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
\STATE Choose $\delta^{(0)} \in (0, 2]$, $\delta_L^{(0)} = 0.99\delta^{(0)}$, $\delta_U^{(0)} = 1.01\delta^{(0)} \wedge 2$.
\FOR{$i=0,1,2, \ldots, N$}
\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_L^{(i)})$ for $j=1, \ldots,M/2$.
\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_U^{(i)})$ for $j=M/2+1, \ldots, M$.
\STATE $\chi(y;\X^{(i)}) = M^{-1}
\left(\sum_{j=1}^{M/2} \nu(y;x_j^{(i)},\delta_L^{(i)})+\sum_{j=M/2+1}^{M} \nu(y;x_j^{(i)},\delta_U^{(i)})\right)$.
\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
\IF{$i$ is in $\{n_k\}_{k=1}$}
	\STATE For $w_{kj} = w_j^{(k)}$, $N$ iterations since last update,
	\STATE $T_L = NM(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj}^2)$.
	\STATE $T_U = NM(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj}^2)$.
	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta\D T$.
\ELSE
	\STATE $\delta^{(i+1)} = \delta^{(i)}$.
\ENDIF
\IF{$i < K$}
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^{M/2} \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=1}^{M/2}$
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=M/2+1}^M \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=M/2+1}^M$
\ELSE
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^M \rightarrow (\frac{1}{M}, x_j^{(i+1)})_{j=1}^M$
\ENDIF
\ENDFOR
\end{algorithmic}
\end{mdframed}\caption{A pseudo-code representation of the adaptive
  PAIS algorithm.}
\label{tab:adapt}
\end{table}
\end{appendix}

>>>>

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins}
\setlength{\parindent}{0in}
\usepackage{amsmath,amsfonts,amsbsy}
\newcommand{\RARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
\newcommand{\RARRlong}[3]{#1
  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
  \longrightarrow}^{#3}\; #2}
\newcommand{\LARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
\newcommand{\bX}{{\bf X}}
\newcommand{\vecx}{{\mathbf x}}
\newcommand{\vecy}{{\mathbf y}}
\newcommand{\vecz}{{\mathbf z}}
\newcommand{\vecq}{{\mathbf q}}
\newcommand{\bs}{{\mathbf s}}
\newcommand{\vecr}{{\mathbf r}}
\newcommand{\vecX}{{\mathbf X}}
\newcommand{\vecv}{{\mathbf v}}
\newcommand{\tick}{\ding{52}}
\newcommand{\cross}{\ding{54}}
\newcommand{\vecn}{{\mathbf n}}
\newcommand{\vecp}{{\mathbf p}}
\newcommand{\cT}{{\mathcal T}}
\newcommand{\dt}{{\mbox{d}t}}
\newcommand{\dx}{{\mbox{d} \vecx}}
\newcommand{\boldnu}{{\boldsymbol \nu}}
\newcommand{\er}{{\mathbb R}}
\renewcommand{\div}{{\rm div}}
\newcommand{\bnu}{{\bf \nu}}
\newcommand{\divergence}{\mathop{\mbox{div}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\FP}{P_{\rm{FP}}}
\newcommand{\ME}{P_{\rm{ME}}}
\newcommand{\MEs}{P_{\rm{ME}_{S}}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\X}{{\mathbf X}}
\newcommand{\Y}{{\mathbf Y}}
\newcommand{\W}{{\mathbf W}}
\newcommand{\data}{D}
\newcommand{\neff}{n_{\text{eff}}}
\newcommand{\E}{{\mathbb E}}

\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are commonly used and powerful
  family of numerical methods for sampling from complex probability
  distributions. As the applications that these methods are being
  applied to increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and with the development of
faster more powerful computers, have become ever more important in a
whole range of fields in statistics, science and engineering. However,
despite the advances in hardware that have made it possible to fully
characteries the posterior distributions of many problems, it remains
unfeasible to use MCMC methods for many more. In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

@@

Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments. A description of the speed up attained by this algorithm is given in Section~\ref{sec:calc_saving}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{"figures/pCNL2o"}
\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL2o"}
\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}. The setup is as in Section~\ref{sec:Implementation P1}.}
\label{fig:P2 opt delta}
\end{figure}

@@

Figure~\ref{fig:MH2 L2} shows that the PAIS-pCNL algorithm, converges to the posterior distribution faster than the pCNL algorithm. The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations.

@@

We can see that the adaptive algorithms compare closely with the
respective nonadaptive algorithms and the improvement PAIS offers
remains significant.

@@

\subsubsection{Calculating the Speed Up in Convergence}\label{sec:calc_saving}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/calc_saving"}
\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2}$ as a percentage of pCNL iterations. The relative L2 error graphs are from the problem in Section~\ref{sec:problem 1}.}
\label{fig:calc_saving}
\end{center}
\end{figure}

The graphs in the previous section clearly show that the PAIS-pCNL algorithm converges faster than the pCNL algorithm. We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving. In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings. We regress through the relative L2 error curve for both algorithms fixing the exponent at $-1/2$, i.e. look for the constants $c_1$ and $c_2$ such that
\[
	\text{pCNL L2 error} = c_1x^{-1/2}, \quad \text{and} \quad \text{PAIS-pCNL L2 error} = c_2x^{-1/2}.
\]
Then, for a chosen tolerance, $c_1x_1^{-1/2} = c_2x_2^{-1/2}$, where $x_1$ and $x_2$ are the number of iterations it takes the pCNL and PAIS-pCNL algorithms to reach a particular tolerance, given by the vertical lines on Figure~\ref{fig:calc_saving}. To find a percentage of iterations required by the PAIS-pCNL algorithm to achieve the same tolerance as the pCNL algorithm, we then use the expression
\[
	\frac{x_2}{x_1} = \left(\frac{c_2}{c_1}\right)^2.
\]
A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}.

\begin{table}[!h]
\centering
\begin{tabular}{|l|r|r|}
\hline
		& Gaussian (low KL div) & Gaussian (high KL div) \\ \hline
	RWMH & 14\% & 14\% \\
	pCN & 33\% & - \\
	MALA & 41\% & 42\% \\
	pCNL & 62\% & 66\% \\ \hline
		& Bimodal (low KL div) & Bimodal (high KL div) \\ \hline
	RWMH & 40\% & 44\% \\
	pCN & 32\% & 4\% \\
	MALA & 36\% & 40\% \\
	pCNL & 56\% & 0.02\% \\ \hline
\end{tabular}
\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective Metropolis algorithms.}
\label{table:calc_savings}
\end{table}

\subsubsection{A Useful Property of the PAIS Algorithm for Multimodal Distributions}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_suction"}
\caption{This figure demonstrates the redistribution property of the PAIS algorithm. Initially there is one chain in the positive mode, and 49 chains in the negative mode.}
\label{fig:BM2_suction}
\end{center}
\end{figure}

@@

\begin{appendix}
\section{The adaptive PAIS algorithm}\label{Sec:App}
\begin{table}[!h]
\begin{mdframed}
\begin{algorithmic}
\STATE $x_j^{(0)} \sim \mu_0$, for $j \in L\cup U$, where $L = \{j\}_{j=1}^{M/2},\ U = \{j\}_{j=M/2+1}^M$.
\STATE Choose $\delta^{(1)} \in (0, 2]$. Set $\delta_{L,U}^{(1)} = (1\pm 0.01)\delta^{(1)}\wedge 2$.
\FOR{$i=1,2, \ldots, N$}
\STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\delta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\delta_U^{(i)})$ for $j \in U$.
\STATE Calculate 
\[
	w^{(i)}_j = \frac{\pi(y_j^{(i)})}{\nu(y_j^{(i)};X^{(i-1)})},
\]
where
\[
	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\delta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\delta_U).
\]

\IF{$i$ is in $\{n_k\}_{k=1}$}
	\STATE For $w_{kj} = w_j^{(k)}$, $S = n_k - n_{k-1}$,
	\STATE $T_L = SM(\sum_{k=i-S}^S\sum_{j\in L} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in L} w_{kj}^2)$.
	\STATE $T_U = SM(\sum_{k=i-S}^S\sum_{j\in U} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in U} w_{kj}^2)$.
	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\delta_U^{(i)}-\delta_L^{(i)}}$.
\ELSE
	\STATE $\delta^{(i+1)} = \delta^{(i)}$.
\ENDIF

\IF{$i < N_\text{stop}$}
	\IF{$i < N_\text{join}$}
		\STATE Resample 
	\[
		(w_j^{(i)},y_j^{(i)})_{j\in L} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in L}, \quad (w_j^{(i)},y_j^{(i)})_{j\in U} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in U}.
\]
	\ELSE
		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
	\ENDIF
	\STATE $\delta_{L,U}^{(i)} = \delta^{(i)} \pm 2\sqrt{2}\delta^{(i)}/\sqrt{NM} \wedge 2$.
\ELSE
		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
		\STATE $\delta_L^{(i+1)} = \delta_U^{(i+1)} = \delta^{(i+1)}$.
\ENDIF
\ENDFOR
\end{algorithmic}
\end{mdframed}\caption{A pseudo-code representation of the adaptive
  PAIS algorithm.}
\label{tab:adapt}
\end{table}
\end{appendix}
